---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
ggplot(lab6_data, aes(x=ndvi_20_med)) +
  geom_histogram() 

ggplot(lab6_data, aes(x=maxtemp_20_med)) +
  geom_histogram() 

ggplot(lab6_data, aes(x=mintemp_20_med)) +
  geom_histogram() 

ggplot(lab6_data, aes(x=rain_20_sum)) +
  geom_histogram() 

ggplot(lab6_data, aes(x=pop_20)) +
  geom_histogram() 

ggplot(lab6_data, aes(x=water_20_pct)) +
  geom_histogram() 

ggplot(lab6_data, aes(x=elev_med)) +
  geom_histogram() 
```

ndvi_20_med does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is right skewed.

maxtemp_20_med does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is left skewed.

mintemp_20_med does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is left skewed.

rain_20_sum does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is right skewed.

pop_20 does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is right skewed.

water_20_pct does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is right skewed.

elev_med does not appear to be normally distributed. The histogram does not resemble a bell curve and the data is right skewed.

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
library(tmap)
#tm_shape(lab6_data) +
#  tm_polygons("ndvi_20_med")


map_ndvi <- qtm(lab6_data, fill="ndvi_20_med", polygons.style = "jenks", polygons.palette ="YlGn", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)
map_maxtemp <- qtm(lab6_data, fill="maxtemp_20_med", polygons.style = "jenks", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)
map_mintemp <- qtm(lab6_data, fill="mintemp_20_med", polygons.style = "jenks", polygons.palette = "-Blues", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)
map_rain <- qtm(lab6_data, fill="rain_20_sum", polygons.style = "jenks", polygons.palette = "Greens", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)
map_pop <- qtm(lab6_data, fill="pop_20", polygons.style = "jenks", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)
map_water <- qtm(lab6_data, fill="water_20_pct", polygons.style = "jenks", polygons.palette = "YlGnBu", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)
map_elevation <- qtm(lab6_data, fill="elev_med", polygons.style = "jenks", polygons.palette = "-plasma", layout.legend.position = c("left", "bottom"), layout.legend.title.size=0.65, layout.legend.width = 1)

tmap_arrange(map_ndvi, map_maxtemp, map_mintemp, map_rain, map_pop, map_water, map_elevation)

```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

The biggest "hotspot" of high NDVI is in Southeast Australia close to the coast. This area has lower median minimum temperatures, more rain, and areas of higher population than most other areas of the country. Southeast Australia also has the highest median elevation of any area in the country. This appears to suggest an association between NDVI and these variables: the more cold, rainy, populous, and elevated an area is, the higher its NDVI/vegetation index is.

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
model_maxtemp<-lm(ndvi_20_med~maxtemp_20_med, data=lab6_data)
model_mintemp<-lm(ndvi_20_med~mintemp_20_med, data=lab6_data)
model_rain<-lm(ndvi_20_med~rain_20_sum, data=lab6_data)
model_pop<-lm(ndvi_20_med~pop_20, data=lab6_data)
model_water<-lm(ndvi_20_med~water_20_pct, data=lab6_data)
model_elev<-lm(ndvi_20_med~elev_med, data=lab6_data)

summary(model_maxtemp)
summary(model_mintemp)
summary(model_rain)
summary(model_pop)
summary(model_water)
summary(model_elev)
```

NDVI and max temp have a statistically significant relationship. For every 1-degree increase in median max temp, the NDVI decreases by 0.0130902. The R-squared is 0.20 suggesting that only a small amount of variance (20%) in NDVI is explained by median max temp.

NDVI and min temp have a statistically significant relationship. For every 1-degree increase in median min temp, the NDVI decreases by 0.012282. The R-squared is only 0.14 suggesting that only a small amount of variance (14%) in NDVI is explained by median max temp.

NDVI and rainfall have a statistically significant relationship. For every 1mm increase in rainfall, the NDVI increases by approx. one-one millionth. The R-squared is 0.42 suggesting that a substantial amount of variance (42%) in NDVI is explained by the amount of rainfall.

NDVI and population have a statistically significant relationship. For every 1 person increase to the population, the NDVI increases by 1.500 * 10^(-6). The R-squared is 0.12 suggesting that only a a small amount of variance (12%) in NDVI is explained by population.

NDVI and water cover do not have a statistically significant relationship. The p-value is 0.249 which is much greater than 0.05.

NDVI and elevation have a statistically significant relationship. For every 1 meter increase in median elevation, the NDVI increases by 1.787 * 10^(-4). The R-squared is 0.05 suggesting that practically none of the variance (5%) in NDVI is explained by elevation.

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
model_multivariate<-lm(ndvi_20_med~mintemp_20_med+rain_20_sum+pop_20+elev_med, data=lab6_data)

summary(model_multivariate)
```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

The coefficient of mintemp_20_med is -1.391e-02. For every one degree increase in min temp, NDVI decreases by 0.01391. This result is similar to that of the univariate model. Min temp and NDVI have a highly statistically significant relationship with an extremely small p value.

The coefficient of rain_20_sum is 9.420e-07. For each mm increase in rainfall, NDVI increases by 9.420e-07. This result is similar to that of the univariate model. Rainfall amount and NDVI have a highly statistically significant relationship with an extremely small p value. 

The coefficient of pop_20 is 2.424e-07. For every 1 person increase in population, NDVI increases by 2.424e-07. The regression coefficient from the univariate model is still extremely small but is about 6 times larger than the population variable's regression coefficient in the multivariate model. Population and NDVI have a statistically significant relationship with a p value of 0.02. 

The coefficient of elev_med is 1.028e-04. For every one meter increase in median elevation, NDVI increases by 1.028e-04. This result is similar to that of the univariate model. Median elevation and NDVI have a highly statistically significant relationship with an extremely small p value.


**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
library(spdep)

lab6_data$residuals <- residuals(model_multivariate)
hist(lab6_data$residuals)

australia_grid_nb <- poly2nb(lab6_data, queen=TRUE)
grid_weight <- nb2listw(australia_grid_nb,style="W",zero.policy=TRUE)

# Moran's I test

moran.test(lab6_data$ndvi_20_med,grid_weight)

moran.mc(lab6_data$ndvi_20_med,grid_weight,nsim=499)

# Getis-Ord Gi*

local_g<-localG(lab6_data$ndvi_20_med,grid_weight)
  
lab6_data$localG<-local_g

# identifies clusters of high and low spatial autocorrelation

lab6_data <- lab6_data %>%
  mutate(localG_recode=case_when(localG <= -1.96~"Cold",
                                 localG < 1.96~"No value",
                                 localG >= 1.96~"Hot"))

tm_shape(lab6_data)+
  tm_polygons("localG_recode")


```

{Summary goes here}

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
#Code goes here
```

{Interpretation goes here}

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

{Discussion here}

**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

{Response here.}

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

{Discussion here.}


**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}

```




#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}

```

